HO unification from object language to meta language

### 2024-02-05 main.tex
print bib
Mv file in elpi_code folder
Update tex

\author{Davide Fissore}
\title{Report - HO unification in Elpi} 
\date{January 29, 2024}

@@

\section{Higher-order unification in \texorpdfstring{$L_\lambda$}{L\_lambda}}

Higher-order unification is a unification problem known to be undecidable. The
goal is to determine a most general unifier between two lambda terms, where at
least one of contain a sub-term $F$ applied to a non empty list of arguments
$a_1,\dots,a_n$ and $F$ is a variable. Some researches have been done in order to
identify some easy patterns where it is possible to solve this kind of
unification.

In \cite{miller1991}, D. Miller provides a unification algorithm (the function
$\eqtau$) for higher-order variable when they are in the so-called pattern
fragment.

\section{From the object to the target language}

Higher order unification becomes difficult when we try to unify terms
represented in the syntax of an object language embedded inside another
language. This kind of interaction between different languages is something that
happens quite often in \coqelpi, a \coq plugin embedding \elpi.

\elpi is a variant of \lambdaprolog equipped with constraint handling rules. On
the other hand, \coq is a proof assistant used to reason about mathematical
properties and certify them thanks to its typing system.

Both languages allows to quantify higher-order variables and are able to perform
higher-order unification when needed. However, when we want to reason with the
\coq unification inside \elpi, it is less evident to reproduce the unification
between two terms. With this paper, we want to provide a framework explaining
how the unification of an object language (\coq in our case) can be expressed in
the targeted language (\elpi in our case).

Without any loss of generality, in \cref{table:coq2elpi} we present a subset of
the terms in the object language and for each term we give its representation in
the targeted language.

\begin{table}
  \centering
  \begin{tabular}{c||c|c}
  Term        & Obj. lang.                    & Targ. lang.                                   \\
  \hline\hline
  Constant    & a                             & `a'                                     \\
  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
  Variable    & X                             & X                                             \\
  \end{tabular}
  \caption{Obj. lang. terms to targ. lang. terms representation}
  \label{table:coq2elpi}
\end{table}

\section{First-order unification}

@@

It is quite evident that a valid substitution for \cref{eq:unif-fo-coq} is $\theta =
\{Y \mapsto `x', Z \mapsto `1'\}$. Now let's consider the same problem translated in
the targeted language.

@@

\section{Higher-order unification}

@@

If we translate the problem before in the target language, the unification
problems showed above become

@@

Now, the new unification problems are no more expressed in the logic of the
targeted language and, therefore, in both cases, unification fails. The
procedure we can adopt in order to transform a higher-order unification problem
of the object language into the logic of the targeted language is to transform
the entry of the problem in a problem which can be understood by the targeted
language. The procedure is made of two steps:

@@

For instance, the former unification problem produce the substitution $\theta_1
= \{F' \mapsto (x\backslash\ f\ x\ 1)\}$. The ho-link function is then applied
to transform the substitution of $F'$ into the corresponding term of the object
langue: $F \mapsto fun \_\ \_\ (x\backslash\ f\ x\ 1)$ which correspond to the
term $fun\ x \Rightarrow f\ x\ 1$. The latter unification problem gives the
substitution $\theta_2 = \{H' \mapsto (y\ x \backslash G'\ x\ y)\}$ in the
targeted language. The first ho-link simply unify $G$ to $G'$ since $G'$ is
flexible, whereas $H$ is mapped to $fun\ y\ x \Rightarrow G'\ x\ y$.

The role of the ho-link is not only to instantiate the higher-order variable F
of the object language when F is flexible and the twin variable in the targeted
language is rigid. It may happen that F has already been partially instantiated.
The unification problem below gives such an example in the object language:

@@

\noindent The first line produces the same substitution as before: $\theta = \{H'
\mapsto (x\ y \backslash G'\ y\ x)\}$, where $G'$ is unified with $G$ and $H$ is
instantiated to <<$fun\ x\ y \Rightarrow G'\ y\ x$>>. While executing the second
line, we see $H''$ instantiated to the function $(x\ y\backslash x)$ in the
targeted language. The last ho-link is charged to link $H''$ with $H$ but this
time $H$ has already been partially instantiated. In particular the call to this
ho-link is as follows: $ho-link\ (x\ y\backslash x)\
(prod\ \_\ \_\ (x\backslash\ prod\ \_\ \_\ (y\backslash\ app [G', x, y])))$.

@@

>>>>

\author{Enrico Tassi, Davide Fissore}
\title{HO unification from object language to meta language} 
\date{Month XX, 202Y}

@@

\begin{abstract}

Unification is a fundamental process behind goal resolution in logic
programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
substitution is a mapping from unification variables to terms and it is the most
general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
Unification has been widely studied in the 90s [here the refs]. In this paper we
want to give a framework to reason about unification in meta-programming.

\end{abstract}

\section{Introduction}

Meta programming \cite{hill1994} is a programming technique in which a program
can treat an other program as its data. This latter program is called object
language, while the former is called meta language. At the heart of meta
programming lays the necessity of representing terms of the object language in
the meta language so that a wide set of program manipulation ranging from
interpretation to compilation. Furthermore, since a logic program is a set of
clauses, the behavior of the object language can be modified by simply adding
new rules inside the database of the meta program.

The motivation pushing us to reason about unification of terms of the object
language represented as terms of the meta language is to implement a type class
solver for the \coq proof assistant into the \elpi logic programming language. A
type class \cite{wadler89, sozeau08} is a typing structure allowing to introduce
ad hoc polymorphism in functional languages and instances are implementations of
type classes. From the point of view of logic programming, type class represent
predicates parametrized by their argument and instances rules for those
predicates. For instance, if \texttt{XXX} is the type class representing the
AAA, then ZZZ and WWW are instances for XXX. In the code snippet below, we give
such implementation in \coq.

\begin{mintedTcolorbox}{coq}
  Class XXX Args := {methods ...}.
  Instance ZZZ : bla bla -> XXX Args := {...} Qed.
  Instance WWW : bla bla -> XXX Args := {...} Qed.
\end{mintedTcolorbox}

This small set of instance thanks a first phase of class and instance\todo{explain compilation of pred and inst?}
compilation is translated into the following \elpi program:

\begin{mintedTcolorbox}{elpi}
  type XXX Args .. -> term -> o.
  XXX Args ... (ZZZ P1 ... PN) :-
    premese...
  WWW Args ... (ZZZ P1 ... PN) :-
    premese...
\end{mintedTcolorbox}

Type-class resolution starts from a query, that is a class applied to some
argument. This \coq term is translated into a term of the meta language and the
search for a solution in the database is started. However, since unification is
performed on the head of the clause, it may happen that the term representation
in the meta language is not sufficiently powerful to understand that two terms
unifying in the object language should also unify in the meta language.

% \section{Higher-order unification in \texorpdfstring{$L_\lambda$}{L\_lambda}}

% Higher-order unification is a unification problem known to be undecidable. The
% goal is to determine a most general unifier between two lambda terms, where at
% least one of contain a sub-term $F$ applied to a non empty list of arguments
% $a_1,\dots,a_n$ and $F$ is a variable. Some researches have been done in order to
% identify some easy patterns where it is possible to solve this kind of
% unification.

% In \cite{miller1991}, D. Miller provides a unification algorithm (the function
% $\eqtau$) for higher-order variable when they are in the so-called pattern
% fragment.

\section{From the object to the meta language}

In \coq, terms unify up to $\eta\beta\delta\iota\zeta$-equivalence and
unification is also possible when dealing with higher order variables. Whereas
\elpi allows higher order unification up to $\eta\beta$-equivalence, and the
representation of \coq terms is more syntactical than one would expect, making 
two terms that unify in \coq not unifiable in \elpi.

Without any loss of generality, in \cref{table:coq2elpi} we present a subset of
the terms in the object language and for each term we give its representation in
the meta language.

\begin{table}
  \centering
  \begin{tabular}{c||c|c}
  Term        & Obj. lang.                    & Meta lang.                                   \\
  \hline\hline
  Constant    & a                             & `a'                                     \\
  Application & f a\_1 a\_2 ... a\_n          & app{[}`f', `a\_1', `a\_2', ..., `a\_n'{]}     \\
  Abstraction & fun (x : T) $\Rightarrow$ f x & fun `x' T (x\textbackslash app{[}`f', `x'{]}) \\
  Variable    & X                             & X                                             \\
  \end{tabular}
  \caption{Obj. lang. terms to meta lang. terms representation}
  \label{table:coq2elpi}
\end{table}

\subsection{First-order unification}

@@

It is quite evident that a valid substitution for \cref{eq:unif-fo-coq} is $\theta =
\{Y \mapsto `x', Z \mapsto `1'\}$. Now let's consider the same problem translated in
the meta language.

@@

\subsection{Higher-order unification}

@@

If we translate the problem before in the meta language, the unification
problems showed above become

@@

Now, the new unification problems are no more expressed in the logic of the
meta language and, therefore, in both cases, unification fails. The
procedure we can adopt in order to transform a higher-order unification problem
of the object language into the logic of the meta language is to transform
the entry of the problem in a problem which can be understood by the meta
language. The procedure is made of two steps:

@@

For instance, the former unification problem produce the substitution $\theta_1
= \{F' \mapsto (x\backslash\ f\ x\ 1)\}$. The ho-link function is then applied
to transform the substitution of $F'$ into the corresponding term of the object
langue: $F \mapsto fun \_\ \_\ (x\backslash\ f\ x\ 1)$ which correspond to the
term $fun\ x \Rightarrow f\ x\ 1$. The latter unification problem gives the
substitution $\theta_2 = \{H' \mapsto (y\ x \backslash G'\ x\ y)\}$ in the
meta language. The first ho-link simply unify $G$ to $G'$ since $G'$ is
flexible, whereas $H$ is mapped to $fun\ y\ x \Rightarrow G'\ x\ y$.

The role of the ho-link is not only to instantiate the higher-order variable F
of the object language when F is flexible and the twin variable in the meta
language is rigid. It may happen that F has already been partially instantiated.
The unification problem below gives such an example in the object language:

@@

\noindent The first line produces the same substitution as before: $\theta = \{H'
\mapsto (x\ y \backslash G'\ y\ x)\}$, where $G'$ is unified with $G$ and $H$ is
instantiated to <<$fun\ x\ y \Rightarrow G'\ y\ x$>>. While executing the second
line, we see $H''$ instantiated to the function $(x\ y\backslash x)$ in the
meta language. The last ho-link is charged to link $H''$ with $H$ but this
time $H$ has already been partially instantiated. In particular the call to this
ho-link is as follows: $ho-link\ (x\ y\backslash x)\
(prod\ \_\ \_\ (x\backslash\ prod\ \_\ \_\ (y\backslash\ app [G', x, y])))$.

@@

\printbibliography
